---
title: "pqcrypto - Post-Quantum Cryptography"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pqcrypto - Post-Quantum Cryptography}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## A brief introduction

Cryptography is the study and practice of techniques to secure communications and data confidentiality, integrity and authenticity through the use of mathematical techniques and algorithms.

**Symmetric key cryptography** uses algorithms that use of the same cryptographic key (aka, symmetric key or secret key) for plain text encryption and for cipher text decryption. This requires that all parties have access to the same key.

**Asymmetric key cryptography**, also known as public-key cryptography, uses pairs of related keys - public key and private key - generated based on 'hard-to-solve' mathematical problems and are widely used for securing communications and data in various applications. 

The rise of quantum computing brings challenges to the security provided by existing asymmetric key algorithms because such computers have the potential to use different algorithms to solve some mathematical problems much faster than classical computers.  
This includes the ability to factor large numbers (e.g. [Shor's Algorithm](https://en.wikipedia.org/wiki/Shor%27s_algorithm)) and compute discrete logarithms, which are fundamental operations in many asymmetric encryption schemes like [RSA](https://en.wikipedia.org/wiki/RSA_(cryptosystem)) or [Diffie-Hellman](https://en.wikipedia.org/wiki/Diffie%E2%80%93Hellman_key_exchange).

The National Institute of Standards and Technology (NIST) has been actively involved in the research and standardization of post-quantum cryptography (PQC) algorithms to address these threats. NIST initiated a public competition known as the "NIST Post-Quantum Cryptography Standardization" to identify and evaluate new cryptographic algorithms that can resist attacks from both classical and quantum computers. This competition is still ongoing and as of the time of this writing (01/Sep/2023), NIST is [requesting comments](https://csrc.nist.gov/news/2023/three-draft-fips-for-post-quantum-cryptography) on 3 drafts for proposed "Federal Information Processing Standards" (FIPS) covering key encapsulation (FIPS-203) and digital signatures (FIPS-204 and FIPS-205).


## The `pqcrypto` package

The `pqcrypto` package makes available to R users some of the latest developments in post-quantum cryptography, covering:

  - Secure key establishment (secret key encapsulation);
  - Digital signatures.


### Secure Key Establishment
*Alice* wants to send *Bob* a secret message. They may meet in person beforehand and agree on a secret key to be used on their communication, but this may not be possible! They need to find a way to exchange such secret key using public channels!
This is possible using key encapsulation techniques. In the example below Bob will generate a key-pair and make his public key available publicly. Alice will use Bob's public key to generate an encapsulation of a secret key to be shared between them. This encapsulated key is only readable by someone in possession of the matching private key. 

```{r key_generation}
library(pqcrypto)

# Bob generates a key-pair with default parameters
bob_keypair <- keygen_kyber()

# Bob makes his public key widely available (e.g. he publishes it on his webpage or blog)
bob_public_key <- bob_keypair$public

# Alice generates a shared secret key and its encapsulation.
result <- encap_kyber(bob_public_key)

# Alice keeps the shared secret for herself and sends the encapsulation to Bob.
alice_shared_secret <- result$shared_secret
send_to_bob <- result$encapsulation

# Bob receives the key encapsulation from Alice and uses his private key to get the shared secret.
bob_shared_secret <- decap_kyber(send_to_bob, bob_keypair$private)

# If everything went well, they both share a secret key. This can be used to apply symmetric 
# encryption (e.g. AES) to a message they want to exchange. 
identical(alice_shared_secret, bob_shared_secret)

```


### Digital Signatures
Digital signatures is a cryptographic technique that can be used to verify a message, document, or transaction __integrity and authenticity__. It provides a way to ensure that the sender of the message is who they claim to be and that the message has not been tampered with during transmission. 
Once the 

##### Example

Alice sent a message to Bob and he wants to make sure the message has not been change. In the example below Alice will generate a key-pair and make her public key publicly available. Alice will send a message and a digital signature generated for that message with her private key and Bob will use Alices's public key to verify the message is authentic.

```{r digital_signatures}
library(pqcrypto)

# Alice generates a key-pair with default parameters
alice_keypair <- keygen_dilithium()

# Alice makes her public key widely available (e.g. he publishes it on her webpage or blog)
alice_public_key <- alice_keypair$public

# Alice writes a message and creates a signature for it.
message_to_bob <- "Hi Bob! Do you want to have some drinks tonight? ;) Alice <3"
message_signature <- sign_dilithium(alice_keypair$private, message_to_bob)

# In a perfect World, Bob would have received the message and the signature from Alice.
# He would have used Alice's public key and checked that the message is authentic. Cheers!
verify_dilithium(message_to_bob, message_signature, alice_public_key)

# But the World is far from perfect!
# Carla, another of Bob's friends who has a crush on him, doesn't want the couple to meet.
# She intercepts Alice's message and change it, but since she does not have access to
# Alice's private key, she cannot forge a new signature. She sends the new message anyway
# in the hope that Bob does not verify it.
message_to_bob <- "Hi Bob. Please meet me at the airport by 8pm! Alice"

# Bob receives the message, presumably from Alice, and checks if it is authentic, to discover
# the message has been tampered.
verify_dilithium(message_to_bob, message_signature, alice_public_key)

```


### NIST's Strength Categories

NIST defines 5 different broad security strength categories as per the table bellow. For an Algorithm to comply with one of them it must be expected that at least the same resources are required to break it as those to preform the corresponding attack type.

| Security Category | Corresponding Attack Type                   | Example  |
|:-----------------:|:--------------------------------------------|:--------:|
|         1         | Key search on block cipher with 128-bit key | AES-128  |
|         2         | Collision search on 256-bit hash function   | SHA3-256 |
|         3         | Key search on block cipher with 192-bit key | AES-192  |
|         4         | Collision search on 384-bit hash function   | SHA3-384 |
|         5         | Key search on block cipher with 256-bit key | AES-256  |
: Table 1 - NIST Security Categories  --  Source: NIST^[National Institute of Standards and Technology (2023) Module-Lattice-Based Digital Signature Standard. (Department of Commerce, Washington, D.C.), Federal Information Processing Standards Publication (FIPS) NIST FIPS 204 ipd. https://doi.org/10.6028/NIST.FIPS.204.ipd]


